\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-05-05 main.tex
cleans typos in first parts

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one's own payoff? In
evolutionary game theory individuals are not required to be rational, instead
they adapt strategies based on mutation and exploration. Strategies are more
likely to spread if they have a high fitness either because the individuals who
adopt them have more offsprings, or because they are imitated more often. The
fitness of a strategy is not constant but depends on the composition of the
population. Individuals interact based on their strategies with other
members of the population and the payoffs they yield are translated into
fitness.

It is commonly assumed that fitness is equivalent to the mean payoff an
individual achieves over the different types in the population, over multiple
interactions. These payoffs assume that individuals can interact with the entire
population several times and remember each and every outcome. Thus, they imply
that individuals have a perfect memory.
However, in their pairwise interactions when they make decisions in each turn
they are assumed to have very limited memory. To be precise most of
the works in the literature,  with a few notable exceptions~\cite{Hauert1997,
Stewart2016}, focus on naive subjects who can only choose from a restricted set
of strategies~\cite{Nowak1992tit}, or who do not remember anything beyond the
outcome of the very last round~\cite{Baek2016}.

The perfect memory assumption is not only unrealistic but it also creates this
curious inconsistency. This has led us to question how robust is our
understanding of cooperation. In this work we explore whether direct
reciprocity can evolve if individuals only remember a minimum of social
information. Though we are not the first to question the assumptions of
estimating fitness~\cite{Roca2006}, we are the first to explore the effect
of payoff memory.

Initially, we consider two extreme scenarios. The first is the classical
scenario of the expected payoffs and the alternative scenario where individuals
update their strategies only based on the very last payoff they obtained. In the
later sections, we allow individuals more memory. More specifically, they can
remember up to two turns and up to two interactions.
We present results on  several social dilemmas which include the prisoner's
dilemma and the donation game, the snowdrift game, the stag hunt game and the
harmony game.

@@

The perfect memory assumption occurs at the updating stage. The learner and the
role model are assumed to interact with a representative sample of the
population, and they remember all interactions they participate in. Thus, they
are updating payoffs are based on the mean payoff they achieved over all the
interactions. These payoffs are refer to as the expected payoffs. We will
compare the expected payoffs to payoffs that are calculated when the role model
and learner do not remember all their interactions. In order to account for the
effect of these different methods we explore the cooperation rate within the
resident population over multiple generations. More details on our methodology
are found in Appendix~\ref{appendix:methods}.

@@

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving
conditional cooperation probabilities $p$ and $q$. The discount factor~$\delta$
is comparably high, thus the opening move \(y\) is a transient effect and has
not effect on the outcome. The left panels corresponds to the standard scenario
considered in the literature. It considers players who use expected payoffs to
update their strategies. The right panel shows the scenario considered herein,
in which players update their strategies based on their last round's payoff. The
top panels assume a benefit \(b\) of 3, whereas the bottom assume a benefit of
10.

@@

We further explore the effect of the benefit,
Figure~\ref{fig:cooperation_rate_over_benefit}. The figure suggests that the
expected payoffs always overestimate cooperation. For the limited memory payoffs
the cooperation rate remains unchanged at approximately 50\% once \(b = 5\). In
specific simulations we have performed, if only the last round payoff is
considered then the highest average cooperation rate that was achieved was 0.51.
In comparison, when the expected payoffs were used the highest average
cooperation rate was 0.97.

@@

Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process where we vary the strength of selection. For
weak selection, \(\beta < 1\), the updating payoffs have no effect. The evolved
population and the average cooperating rate are the same for both approaches.
However for strong selection, it can be seen that the cooperating rate increases
with \(beta\), when expected payoffs are used. However, when limited memory
payoffs are used. the cooperating rate decreases.

>>>>

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one’s own? In evolutionary
game theory, individuals are not required to be rational, instead they adapt
strategies based on mutation and exploration. Strategies are more likely to
spread if they have a high fitness either because the individuals who adopt them
have more offspring, or because they are imitated more often. The fitness of a
strategy is not constant but depends on the composition of the population.
Individuals interact based on their strategies with other members of the
population and the payoffs they yield are translated into fitness.

It is commonly assumed that fitness is equivalent to the mean distribution of
types in the population. These payoffs assume that individuals can interact with
the entire population several times and remember each and every outcome. Thus,
they imply that individuals have a perfect memory. However, when they make
decisions in each turn they are assumed to have very limited memory. To be
precise, most of the works in the literature focus on naive subjects who can
only choose from a restricted set of strategies~\cite{Nowak1992tit}, or who do
not remember anything beyond the outcome of the very last round~\cite{Baek2016}.
Note that there are a few notable exceptions~\cite{Hauert1997, Stewart2016}.

The perfect memory assumption is not only unrealistic but it also creates this
curious inconsistency. This has led us to question how robust is our
understanding of cooperation. In this work, we explore whether direct
reciprocity can evolve if individuals only remember a minimum of social
information. Though we are not the first to question the assumptions of
estimating fitness~\cite{Roca2006}, we are the first to explore the effect of
payoff memory.

Initially, we consider two extreme scenarios. The first is the classical
scenario of the expected payoffs and the alternative scenario where individuals
update their strategies only based on the very last payoff they obtained. In the
later sections, we allow individuals more memory. More specifically, they can
remember up to two turns and up to two interactions. We present results on
several social dilemmas which include the prisoner’s dilemma and the donation
game, the snowdrift game, the stag hunt game and the harmony game.

@@

The perfect memory assumption occurs at the updating stage. The learner and the
role model are assumed to interact with a representative sample of the
population, and they remember all interactions they participate in. Thus, their
updating payoffs are based on the mean payoff they achieved over all the
interactions. These payoffs are referred to as the expected payoffs. We will
compare the expected payoffs to payoffs that are calculated when the role model
and learner do not remember all of their interactions. In order to account for
the effect of these different methods, we explore the cooperation rate within
the resident population over multiple generations. More details on our
methodology are found in Appendix~\ref{appendix:methods}.

@@

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving
conditional cooperation probabilities $p$ and $q$. The discount factor~$\delta$
is comparably high, thus the opening move \(y\) is a transient
effect and has no effect on the outcome. The left panel corresponds to the
standard scenario considered in the literature. It considers players who use
expected payoffs to update their strategies. The right panel shows the scenario
considered herein, in which players update their strategies based on their last
round’s payoff. The top panels assume a benefit \(b\) of 3, whereas the bottom
assume a benefit of 10.

@@

We further explore the effect of the benefit,
Figure~\ref{fig:cooperation_rate_over_benefit}. The figure suggests that the
expected payoffs always overestimate cooperation. For the limited memory
payoffs, the cooperation rate remains unchanged at approximately 50\% once \(b =
5\). The highest cooperation rate that was achieved was 0.51 when  the last
round payoff is considered. In comparison, when the expected payoffs were used
the highest average cooperation rate was 0.97.

@@

Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process where we vary the strength of selection. For
weak selection, \(\beta < 1\), the updating payoffs have no effect. The evolved
population and the average cooperating rate are the same for both approaches.
However for strong selection, it can be seen that the cooperating rate increases
with \(beta\), when expected payoffs are used. However, when limited memory
payoffs are used the cooperating rate decreases.
